---
title: R-Lab for BSS
---

## Simulate Sources

Sources are generated from `arima.sim` with different parameters set. The source signals should be very closed to independent and stationary. 

Though not compulsory, the sources are scales to zero-mean and unit-variance before mixing.

This step generate a list of `sim` that includes `x.simple` as normal mixing, and `x.tv` with a time-varying factor $\boldsymbol{\mathcal E}$.

Note that the matrix are transposed from the literature, but is consistent with `ts` class.

```{r source}
N <- 1e4
sim <- list()

omega <- matrix(rnorm(9) , ncol = 3)
epsilon <- matrix(rnorm(9) * 1e-4, ncol = 3)
s1 <- arima.sim(list(ar=c(0.3,0.6)),N)
s2 <- arima.sim(list(ma=c(-0.3,0.3)),N)
s3 <- arima.sim(list(ar=c(-0.8,0.1)),N)
s <- cbind(s1,s2,s3)
s <- apply(s, 2, scale)

sim$x.simple <- s %*% t(omega)
sim$x.tv <- matrix(nrow = nrow(s), ncol = ncol(s))
for (i in 1:N) sim$x.tv[i,] <- s[i,] %*% t(omega) %*% t(diag(3) + i * t(epsilon))

remove(s1,s2,s3,s) #clean
plot.ts(sim$x.tv)
```

In addtion, one parameter (+ the mixed signals) is required for TV-SOBI Algorithm.

```{r param}
lag.max <- 4
X <- sim$x.tv
p <- ncol(X); n <- nrow(X) # dims from X, not params
```

## Step 1. Decompose Autocovariance into 3 parts.

Through a linear estimation method, the autocovariance matrices (up to required lag.max) are decomposed into 3 parts, $Ra$, $Rb$ and $Rc$

The `lm` function is preferred over matricies calculation, which is mainly due to the "computationally singular" problem may persist for inverse-matrix.

```{r step1}
Ra <- Rb <- Rc <- array(dim = c(p, p, lag.max+1))

# elment wise estimation
for (lag in 0:lag.max) {
  for (i in 1:p){
    for(j in 1:p){
      # lag = 1; i = 2; j = 3
      y.design <- X[(1+lag):n,i] * X[1:(n-lag),j] # empirical autocovariance
      x.design <- cbind(rep(1, n-lag), 1:(n-lag), (1:(n-lag))^2)
      est <- lm(y.design ~ x.design - 1)$coefficients # using lm avoids the singular problem
      Ra[i, j, lag+1] <- est[1]
      Rb[i, j, lag+1] <- est[2]
      Rc[i, j, lag+1] <- est[3]
    }
  }
}

# should be symmetric -> fix it
for (i in 1:(lag.max+1)) {
  Ra[,,i] <- (Ra[,,i] + t(Ra[,,i]))/2
  Rb[,,i] <- (Rb[,,i] + t(Rb[,,i]))/2
  Rc[,,i] <- (Rc[,,i] + t(Rc[,,i]))/2
}
```

It can be observed that $Ra > Rb > Rc$ and the last one is really small

## Step 2: JD on $Ra$

This joint diagnolization problem requires first whitening the autocovariance matricies using covariance matrix, and a handy function is created to achieve this. The function itslef is in fact the SOBI algorithm without calculation of (auto)covariance matrix. However, practically, it seems differ from `JADE::SOBI` slightly, which is most likely due to the matrix-eigen problem.

$\boldsymbol W := \boldsymbol\Omega_0 ^{-1}$

```{r step2}
myCovJD <- function(covs){

    require(JADE)
  
  # whitening using S_0^{-1/2}, through eigen calculation
  eig <- eigen(covs[,,1])
  white <- solve(eig$vectors %*% sqrt(diag(eig$values)))
  
  # whiten and format to rjd
  covs.white <- array(dim = c(dim(white),lag.max))
  for (i in 1:lag.max) covs.white[,,i] <- white %*% covs[,,i+1] %*% t(white)
  
  # joint diagnolization for V | note for the transpose of V
  jd <- frjd(covs.white, maxiter = 1e5)
  
  # D is the estimated diagnals
  list(W = t(jd$V) %*% white, D = jd$D, white = white)
}

# now it is easy
jd <- myCovJD(Ra)
W.est <- jd$W
omega.est <- solve(W.est)
```

## Step 3: Look for $\boldsymbol{\mathcal E}$

### Alternative 1: use $Rb$

It is again a linear optimization problem given nicely designed hat-matrix

```{r step3a}
Q <- Ra

# design matricies
H1 <- H2 <- array(0, dim = c(p^2, p^2, lag.max + 1))
for(lag in 0:lag.max){
  for(i in 1:p^2){
      # the column to use  
    Qi <- Q[,ceiling(i/p),lag+1]
      # H1 similar to diagonal
    pos <- (ifelse(i%%p == 0, p, i%%p ) - 1) * p  + (1:p)
    H1[i, pos, lag+1] <- Qi
      # H2 similar to vec
    pos <- (ceiling(i/p) - 1) * p + (1:p)
    H2[i, pos, lag+1] <- Qi
  }
}

y.design <-matrix(as.vector(Rb[,,1]), nrow = p^2)
x.design <- H1[,,1] + H2[,,1]
for (i in 2:p) {
  y.design <- y.design + matrix(as.vector(Rb[,,i]), nrow = p^2)
  x.design <- x.design + H1[,,i] + H2[,,i]
}

# estimate
est <- lm(y.design ~ x.design - 1)$coefficients
epsilon.est1 <- matrix(est, nrow = p)
```

### Alternative 2: use $ \boldsymbol{\hat\Omega \hat\Lambda \hat\Omega}$

The algorithm is exact with the first alternative, but choose $Q$ different by using the result from Step 2.

In fact, if the joint diagnolization is perfect, above two alternatives are exact.

```{r step3b, eval=FALSE}
# not run
# replace the Q in above with
Q <- array(dim = c(p, p, lag.max + 1))
Q[,,1] <- omega.est %*% t(omega.est)
for (lag in 1:lag.max) Q[,,lag + 1] <- omega.est %*% jd$D[,,lag] %*%  t(omega.est)
```

### Alternative 3： Joint Diagnolization on $Rc$

Though $Rc$ is rather small in value, it is possible to estimate via joint diagonlization (for $\boldsymbol{\mathcal E \Omega} $). The accuracy/loss is dependent on the linear estimators in step 1 and the loss from Joint Diagnolization.

```{r step3c}
jd <- myCovJD(Rc)
epsilon.est2 <- solve(jd$W) %*% W.est
```

## Brief Evaluation on the result

```{r md, results=`hold`}
cat("Accuracy of Omega:\n using TV-SOBI", MD(W.est, omega), "\n using SOBI", MD(SOBI(X)$W, omega), "\n\n")
print(epsilon)
print(epsilon.est1)
print(epsilon.est2)
```

## Notes and Further Considerations

The autocovariance matricies are always arranged as in `JADE` package ($dim=c(p,p,lag)$), but is different from `acf` function ($dim=c(lag,p,p)$). Furthermore, the index is always starting with 1, implying that lag=0 is stored in $[,,1]$, and lag=1 in $[,,2]$..

Implementation to class `bss` not yet achieved.

For the compatibility with `JADE` packages, the function `JADE:bsscomponents` may need to be extend in order to correctly restore sources from TV-SOBI.

## Memo on SOBI with new function

```{r sobi}
ans2 <- SOBI(X)
```